Pythran: Bringing OpenMP to a Python Subset

### 2013-02-02 doc/papers/europar2013/paper.tex
Few cosmetic modifications.

OpenMP directives are held by C/C++ pragma, or by Fortran comments. Most of them apply
on structured blocks, represented by an instruction inC/C++ and delimited by
comment in Fortran. A few directives (e.g.\ \texttt{threadprivate}) are not
attached to a specific instruction.

@@

\begin{figure}
    \begin{lstlisting}[label={lst:instruction-annotation}, caption={Parallel
    loop with reduction annotated with OpenMP directive.}]
'omp parallel for reduction(+:r)'
for auto in my_container:
  r+=x
  \end{lstlisting}
\end{figure}

@@

The \texttt{reduction(\emph{operator}: \emph{list}} is used to handle data
dependencies when performing a parallel reduction. The supported operator
depends on the input and backend language: Python has \texttt{min}/\texttt{max} operators but
C/C++ have not. Python does not have \texttt{\&\&} or \texttt{||}
operators~\footnote{Although similar, the \texttt{and} and \texttt{or} keywords
have a special meaning in Python.} while C/C++ have.

@@

The core directive to handle data parallelism is the \texttt{for} directive that
distribute the iteration space of the associated loop among the existing
threads. To be compatible with OpenMP, the loop iteration space must be
described by a random access iterator with a total order, that is it is possible
to increase the iterator by any integer and to decide which iterator is the
greatest among two. The integer used as loop indices in Fortran and C satisfy
this conditions, so do C++ iterators with the \texttt{random\_access\_tag}
trait. But a Python iterator only advance by a step of one until it is
exhausted, in which case it raises an exception. They only have forward iterator
behavior.

@@

Python also generalizes the concept of iterators through generators, which also
behave like forward iterators. The simplest one, \texttt{xrange(\emph{start},
\emph{stop}, \emph{step})}, successively yields value starting from
\texttt\emph{start} to \texttt\emph{stop} by a step of \texttt\emph{step}, is
easily extended to support random access, but it is generally does not make
sense to use a generator as a loop iterator, as the relation between two
random states of the iterator may be of an arbitrary complexity.

@@

\begin{figure}
    \begin{lstlisting}[label={lst:complex-omp-for}, caption={Parallel
    loop with tuple unpacking.}]
'omp parallel for'
for i,v in enumerate([2, 3, 5, 7, 11]):
  print i, ':', v
  \end{lstlisting}
\end{figure}

@@

The \texttt{\_OPENMP} macro definition is always provided when OpenMP is
activated, and can be used for portability with system where OpenMP is not
available. Python does not have a preprocessor, but is possible to catch the
import exception if the \texttt{omp} module is not found.

@@

\begin{figure}
    \begin{lstlisting}[label={lst:openmp-validation},caption={Example of Python
    OpenMP validation test case.}]
def omp_parallel_for_if():
  import omp
  using = num_threads = 0
  sum = sum2 = 0
  LOOPCOUNT=1000
  'omp parallel for private(i) if(using == 1)'
  for i in range(LOOPCOUNT+1):
    num_threads = omp.get_num_threads()
    sum+=i
  known_sum = (LOOPCOUNT * (LOOPCOUNT + 1)) / 2
  return known_sum == sum and num_threads == 1
    \end{lstlisting}
\end{figure}

We have used the Pythran tool described in the following section to turn each
Python test function into a C++ function with the same directives and runtime
calls. Apart from the \texttt{threadprivate} directive and the
\texttt{collapse(\emph{n}} clause, all tests were successful.
\texttt{threadprivate} directives were held by global variables not supported in
Pythran~; and the C++ code generated by Pythran does not preserve the perfect
loop nesting required by the \texttt{collapse} clause.

@@

    \begin{lstlisting}[label={lst:cython-sample}, caption={Cython Impelmentation
    of a Parallel Reduction.}]
import math
cimport cython
from cython.parallel import parallel, prange
from libc.math cimport sqrt
def sum_sqrt():
    cdef float r=0
    cdef int i
    with nogil, parallel():
        for i in prange(10000000):
            r+= sqrt(i)
    return r
\end{lstlisting}
\end{figure}
%
\begin{figure}[ht]
    \begin{lstlisting}[label={lst:pythran-sample}, caption={Cython Impelmentation
    of a Parallel Reduction.}]
#pythran export sum_sqrt()
import math
def sum_sqrt():
    r=0
    "omp parallel for private(i) reduction(+:r)"
    for i in xrange(10000000):
        r+= math.sqrt(i)
    return r
    \end{lstlisting}
\end{figure}

@@

>>>>

OpenMP directives are held by C/C++ pragma, or by Fortran comments. Most of them apply
on structured blocks, represented by an instruction in C/C++ and delimited by
comment in Fortran. A few directives (e.g.\ \texttt{threadprivate}) are not
attached to a specific instruction.

@@

\begin{figure}
    \begin{lstlisting}[label={lst:instruction-annotation}, caption={Parallel
    loop with reduction annotated with OpenMP directive.}]
'omp parallel for reduction(+:r)'
for auto in my_container:
  r += x
  \end{lstlisting}
\end{figure}

@@

The \texttt{reduction(\emph{operator}: \emph{list})} is used to handle data
dependencies when performing a parallel reduction. The supported operator
depends on the input and backend language: Python has \texttt{min}/\texttt{max} operators but
C/C++ have not. Python does not have \texttt{\&\&} or \texttt{||}
operators~\footnote{Although similar, the \texttt{and} and \texttt{or} keywords
have a special meaning in Python.} while C/C++ have.

@@

The core directive to handle data parallelism is the \texttt{for} directive that
distribute the iteration space of the associated loop among the existing
threads. To be compatible with OpenMP, the loop iteration space must be
described by a random access iterator with a total order, i.e.\ it is possible
to increase the iterator by any integer and to decide which iterator is the
greatest among two. The integer used as loop indices in Fortran and C satisfy
this conditions, so do C++ iterators with the \texttt{random\_access\_tag}
trait. But a Python iterator only advance by a step of one until it is
exhausted, in which case it raises an exception. They only have forward iterator
behavior.

@@

Python also generalizes the concept of iterators through generators, which also
behave like forward iterators. The simplest one, \texttt{xrange(\emph{start},
\emph{stop}, \emph{step})}, successively yields value starting from
\texttt\emph{start} to \texttt\emph{stop} by a step of \texttt\emph{step}. It is
easily extended to support random access, but it generally does not make
sense to use a generator as a loop iterator, as the relation between two
random states of the iterator may be of an arbitrary complexity.

@@

\begin{figure}
    \begin{lstlisting}[label={lst:complex-omp-for}, caption={Parallel
    loop with tuple unpacking.}]
'omp parallel for'
for i, v in enumerate([2, 3, 5, 7, 11]):
  print i, ':', v
  \end{lstlisting}
\end{figure}

@@

The \texttt{\_OPENMP} macro definition is always provided when OpenMP is
activated, and can be used for portability with system where OpenMP is not
available. Python does not have a preprocessor, but it is possible to catch the
import exception if the \texttt{omp} module is not found.

@@

\begin{figure}
    \begin{lstlisting}[label={lst:openmp-validation},caption={Example of Python
    OpenMP validation test case.}]
def omp_parallel_for_if():
  import omp
  using = num_threads = 0
  sum = sum2 = 0
  LOOPCOUNT = 1000
  'omp parallel for private(i) if(using == 1)'
  for i in range(LOOPCOUNT + 1):
    num_threads = omp.get_num_threads()
    sum += i
  known_sum = (LOOPCOUNT * (LOOPCOUNT + 1)) / 2
  return known_sum == sum and num_threads == 1
    \end{lstlisting}
\end{figure}

We have used the Pythran tool described in the following section to turn each
Python test function into a C++ function with the same directives and runtime
calls. Apart from the \texttt{threadprivate} directive and the
\texttt{collapse(\emph{n})} clause, all tests were successful.
\texttt{threadprivate} directives were held by global variables not supported in
Pythran yet~; and the C++ code generated by Pythran does not preserve the perfect
loop nesting required by the \texttt{collapse} clause.

@@

    \begin{lstlisting}[label={lst:cython-sample}, caption={Cython Impelmentation
    of a Parallel Reduction.}]
import math
cimport cython
from cython.parallel import parallel, prange
from libc.math cimport sqrt
def sum_sqrt():
    cdef float r = 0
    cdef int i
    with nogil, parallel():
        for i in prange(10000000):
            r += sqrt(i)
    return r
\end{lstlisting}
\end{figure}
%
\begin{figure}[ht]
    \begin{lstlisting}[label={lst:pythran-sample}, caption={Cython Impelmentation
    of a Parallel Reduction.}]
#pythran export sum_sqrt()
import math
def sum_sqrt():
    r = 0
    "omp parallel for private(i) reduction(+:r)"
    for i in xrange(10000000):
        r += math.sqrt(i)
    return r
    \end{lstlisting}
\end{figure}

@@

All the source codes used for these benchmarks are availabe on the Pythran
reporsitory.~\footnote{\emph{cf}.\
\texttt{https://github.com/serge-sans-paille/pythran} on \texttt{europar2013}
branch.}
